---
title: "Bile Acid Based Alzheimer's Prediction - Data Cleaning"
author: "Shadi Saee"
date: "27 - 09 - 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Loading Packages
```{r load packages, warning = FALSE, message = FALSE}
#install.packages(corrplot)
#install.packages("table1")
#install.packages("Hmisc")  #label function for tables, sig values for corrs
#install.packages("splitstackshape")  stratifiation
#install.packages("caret")

library(dplyr)
library(corrplot)
library(table1)
library(Hmisc)
library(stats)
library(caret)
library(splitstackshape)
```


# Loading Data 

```{r load data}
adnimerge <- read.delim("Data/ADNIMERGE.csv", sep = ",", stringsAsFactors = FALSE)
vitals <- read.delim("Data/VITALS.csv", sep = ",", stringsAsFactors = FALSE)
postproc <- read.delim("Biospecimen_Results/ADMC_BA_POSTPROC_06_28_18.csv", 
              sep = ",", stringsAsFactors = FALSE)

```


# Data Preparation and Cleaning: Subsetting Relevant Variables and Instances

## Selecting Variables of Interest

All the loaded datasets contain variables that are irrelevant for the current 
task. I subset only the relevant variables. The adnimerge dataset contains 
many key variables like demographic information, diagnosis, biomarkers, and neuropsychological test results. I select diagnosis (+ baseline diagnosis which has a seperate variable), age, sex, education, APOE4 genotype, and biomarker information at baseline. The vitals dataset contains information on weight and height of the participants neccessary to create the BMI variable. The postproc dataset contains the log-transformed BA measurements taken at baseline and information about the extraction process. I only keep the measurements of the   BAs and their respective ratios. In all datasets I also keep the unique identifier RID and the "VISCODE" variable that encodes which visit the instance belongs too. Each visit entails a new instance. 

I also combine the diagnoses late MCI and early MCI to "MCI" and
turn the sex variable into a factor, as it was previously encoded as 
character.

&nbsp;
```{r Discard irrelevant variables}
# Discard irrelevant variables 
vitals2 <- vitals %>% select(RID, VISCODE, VSWEIGHT, VSWTUNIT, VSHEIGHT, VSHTUNIT)
postproc2 <- postproc[ , -c(1,5:25,34:57)] 
adnimerge2 <- adnimerge[ , c(1,3,8:11,15,60,66,109:110,112)] 
```

```{r Change names/values of some variables}
adnimerge2 <- adnimerge2 %>%
  mutate(DX_bl = ifelse(DX_bl == "LMCI"|DX_bl == "EMCI", 'MCI', DX_bl)) %>%
  mutate(PTGENDER = factor(PTGENDER), APOE4 = factor(APOE4)) %>%
  rename("SEX" = PTGENDER, "EDUCATION" = PTEDUCAT, "ADAS13" = ADAS13_bl) %>% 
  arrange(RID, VISCODE)

```


## Filtering patients who converted within three years 

I subset and manipulate the datasets individually at first and merge them 
together later. 

I exclude patients with subjective memory complaints and include only patients 
who have converted to AD within three years. I do this in two steps:
first I filter for patients that have diagnostic information available at three 
years. In a second step I check for patients who converted from MCI to AD earlier than three years after BL. 


```{r, Filtering for instances of interst}
# Remove patients with SMC and include only patients with data available for 
#three years
adnimerge3 <- adnimerge2 %>% 
  filter(DX_bl == 'MCI')  %>% 
  filter(Years_bl <= 3.2) %>% 
  group_by(RID) %>% 
  mutate(max_data = (max(M))) %>%
  filter(max_data == 36) %>%
  select(-c(max_data, Years_bl)) %>%
  ungroup(RID)

# Also include MCI patients who converted to dementia earlier than three years:
#First, get all RIDs of people who converted from MCI to dementia within three #years. Then, filter the whole dataset for those RIDs and then exclude those RIDs
# who were already in the adnimerge3 object

`%not in%` <- function (x, table) is.na(match(x, table, nomatch = NA_integer_))
# Source: https://stackoverflow.com/questions/5831794/opposite-of-in

MCI_converters_RID <- adnimerge2 %>%
  filter(DX_bl == "MCI") %>%
  filter(Years_bl <= 3.2) %>% 
  filter(DX == "Dementia") %>%
  pull(RID)

MCI_converters <- adnimerge2 %>% 
  filter(RID %in% MCI_converters_RID) %>%
  filter(RID %not in% adnimerge3$RID) %>%
  select(-Years_bl)


# Function to count number of unique participants in dataset:
count_participants <- function(df){
  return (length(unique(df$RID)))
}

# Check how many patients are in the adnimerge3 subset
count_participants(adnimerge3)

# Check  how many patients are in the MCI_converters that were excluded in adnimerge3
count_participants(MCI_converters)

#Check if both dataframes have same variables to be able to append them
all_equal(names(adnimerge3),names(MCI_converters))

# Append MCI_converters to adnimerge3
adnimerge4 <- rbind(adnimerge3, MCI_converters)

count_participants(adnimerge4)      
```

## Create BMI Variable

In the next step I process the information from the vitals table to create 
the BMI variable. This entails first changing all measurements to the same scale (some measurements are in cm and others in inches) and then creating the BMI variable based on weight and height. Three instances were mistakenly coded as inches when they were actually already in cm, resulting in heights above two meters fifty, therefore I mutated those back to the original value.
&nbsp;
```{r BMI Variable}
# Process info for BMI meausure 
bmi <- vitals2 %>% 
  select(RID, VISCODE,VSHEIGHT, VSHTUNIT, VSWEIGHT, VSWTUNIT) %>%
  mutate(VSHEIGHT = ifelse(VSHTUNIT == 1, VSHEIGHT * 2.54, VSHEIGHT)) %>%
  mutate(VSWEIGHT = ifelse(VSWTUNIT == 1, VSWEIGHT / 2.205, VSWEIGHT)) %>%
  filter(VSHEIGHT != -4 & !is.na(VSHEIGHT)) %>%
  filter(VISCODE %in% c("bl", "sc", "v01")) %>%
  arrange(RID)

# Check if heights are correct
too_tall <- bmi[bmi$VSHEIGHT > 200,] 
too_tall

# Convert 3 instances back to original value
too_tall_ids <- too_tall[3:5,1]

mutate_cond <- function(.data, condition, ..., envir = parent.frame()) {
  condition <- eval(substitute(condition), .data, envir)
  .data[condition, ] <- .data[condition, ] %>% mutate(...)
  .data
}
# Source: https://stackoverflow.com/questions/34096162/dplyr-
#mutate-replace-several-columns-on-a-subset-of-rows
  
bmi <- bmi %>%
  mutate_cond(RID %in% too_tall_ids, VSHEIGHT = VSHEIGHT/2.54) 
bmi[bmi$RID %in% too_tall_ids,]


bmi <- bmi %>% 
  mutate(VSHEIGHT = VSHEIGHT/100) %>%
  mutate(BMI = VSWEIGHT/(VSHEIGHT*VSHEIGHT)) %>% 
  mutate(VISCODE = "bl") %>%
  select(-c(VSHTUNIT,VSWTUNIT))

```

## Merging datasets

In the next step I merge all three datasets via left join and then subset the 
merged dataset to include only participants with bile acid meaurements available.

```{r Merge data}
# Merge data 
total <- adnimerge4 %>%
  left_join(bmi, by = c("RID", "VISCODE")) %>%
  left_join(postproc2, by = c("RID", "VISCODE")) %>% 
  arrange(RID, VISCODE)
  
count_participants(total) #632

# Only participants with BA info available 
final <- total[total$RID %in% postproc2$RID,]

count_participants(final)  #609
```

## Create Conversion Variable

Next I create a categorical conversion variable that enocodes whether the 
patients converted from MCI to AD within three years or not. First, I create 
a new variable "Convert" that has 0 as value for everyone. Then I create a function that changes all values of the "Convert"variable for each patient to 1  if the diagnosis for that patients equals "Dementia" at any point in time. 

```{r Create Conversion Variable}
final <- mutate(final, CONVERT = 0)

create_convert <- function(df){
  for (i in unique(df$RID)){
    if (any(df[df$RID == i,]$DX == "Dementia")){
      df$CONVERT[df$RID == i] <- 1
    }
  }
  df
}

final <- create_convert(final) 
final$CONVERT <- factor(final$CONVERT)
```

```{r Function to check conversion variable, eval = FALSE}
for(i in unique(final$RID)){
  subgroup <-  filter(final, RID == i)
  print(subgroup %>% select(RID, VISCODE, DX, CONVERT))
}

final %>% group_by(CONVERT) %>% summarise(n_distinct(RID)) %>% ungroup()
```

Now, all the information we need is encoded at baseline visit. Thus, I can subset the MCI patients object to only include the baseline visit and delete some superfuous variables

```{r MCI Baseline visit subset}
MCIBL <- final %>% 
  filter(VISCODE == "bl") %>%
  select(-c(VISCODE,DX_bl,DX, Month_bl,M,PLATFORM_ID)) %>%
  select(RID:ADAS13, BMI, CONVERT, CA_CDCA:TDCA_DCA) %>%
  as.data.frame()

write.csv(MCIBL, "MCIBL.csv")
```

```{r Create data partion}
set.seed(7)
trainIndex <- createDataPartition(MCIBL$CONVERT, p = .85, 
                                  list = FALSE)


